Recurrent state space models capture discrete and continuous dynamics of neural activity in \textit{C. elegans}

### 2018-08-29 drafts/elife/draft.tex
adding hierarchical ar observations

% \title{\large Discrete and continuous latent states of neural activity in \textit{Caenorhabditis elegans}}
% \title{Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}}
\title{Hierarchical state space models reveal dynamics of neural activity in \textit{C. elegans}}
\author{Scott Linderman$^{\text{1}}$,
  Annika Nichols$^{\text{2}}$,
  David Blei$^{\text{1}}$,
  Manuel Zimmer$^{\text{2}}$,
  and
  Liam Paninski$^{\text{1}}$
  \\
  $^{\text{1}}$Columbia University,
  $^{\text{2}}$Research Institute of Molecular Pathology (IMP), Vienna Biocenter
}

@@

We seek a parsimonious characterization of the dynamics of neural
activity in the head ganglia of \celegans. Recent advances in optical
imaging offer simultaneous recordings of most of the head ganglia
neurons, but we must surmount three main challenges in order to learn
a dynamics model from these data.  First, we need to design a model
flexible enough to capture nonlinear dynamics yet interpretable enough
to reveal simplifying structure when it exists.  Moreover, any model
will necessarily be an approximation to the true neural dynamics, and
as such we need to be robust to data that does not quite match model
predictions.  Finally, our model must combine partial recordings from
separate organisms to learn shared neural dynamics yet still allow for
individual variability. We construct a recurrent, robust, and
hierarchical state space model that addresses these challenges,
decomposing complex dynamics into a collection of simpler pieces that
can be learned from noisy, partial recordings with trial-to-trial
variability.  We call this model \textbf{TBD}.\todo{name it!}

@@

These activations lend further evidence in support of the behavioral
labels attributed to each state. For eaxample, state 4 (green)
coincides with the manually labeled ``ventral turn'' state.  Its
dynamics produce a pronounced rise and fall in the activity of cluster 6, which
(referring back to Fig.~\ref{fig:clustering}), consists of the
\textsf{RIVL/R} and \textsf{SMDVL/R} neurons.  The \textsf{RIV} pair
are polymodal inter- and motor-neurons that specify the ventral bias
of turns, and the \textsf{SMDV} are motor neurons that initiate
ventral turns.  We also see that states 6 (orange) and 5 (purple) often
produce a dip followed by a rise, respectively, of activation on cluster 8,
which together produce a dorsal turn.
\todo{Need confirmation of these results.  Hard for me to tell difference between
dorsal turn activation (orange to purple) and ventral turn (green to brown).}

@@

\paragraph{Input-driven transition probabilities}
todo
\todo[inline]{Detail the model for how transition probabilities depend
  on oxygen and change in oxygen concentration.  Straightforward extension
  of the recurrent model above.}

>>>>

% \title{\large Discrete and continuous latent states of neural activity in \textit{Caenorhabditis elegans}}
% \title{Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}}
\title{Recurrent state space models capture discrete and continuous dynamics of neural activity in \textit{C. elegans}}
\author{Scott Linderman$^{\text{1}}$,
  Annika Nichols$^{\text{2}}$,
  David Blei$^{\text{1}}$,
  Manuel Zimmer$^{\text{2}}$,
  and
  Liam Paninski$^{\text{1}}$
  \\
  $^{\text{1}}$Columbia University,
  $^{\text{2}}$Research Institute of Molecular Pathology (IMP), Vienna Biocenter
}

@@

We seek a parsimonious characterization of the dynamics of neural
activity in the head ganglia of \celegans. Recent advances in optical
imaging offer simultaneous recordings of most of the head ganglia
neurons, but we must surmount three main challenges in order to learn
a dynamics model from these data.  First, we need to design a model
flexible enough to capture nonlinear dynamics yet interpretable enough
to reveal simplifying structure when it exists.  Moreover, any model
will necessarily be an approximation to the true neural dynamics, and
as such we need to be robust to data that does not quite match model
predictions.  Finally, our model must combine partial recordings from
separate organisms to learn shared neural dynamics yet still allow for
individual variability. We construct a recurrent, robust, and
hierarchical state space model that addresses these challenges,
decomposing complex dynamics into a collection of simpler pieces that
can be learned from noisy, partial recordings with trial-to-trial
variability.  We call this model \textbf{TBD}.
%\todo{name it!}

@@

These activations lend further evidence in support of the behavioral
labels attributed to each state. For eaxample, state 4 (green)
coincides with the manually labeled ``ventral turn'' state.  Its
dynamics produce a pronounced rise and fall in the activity of cluster 6, which
(referring back to Fig.~\ref{fig:clustering}), consists of the
\textsf{RIVL/R} and \textsf{SMDVL/R} neurons.  The \textsf{RIV} pair
are polymodal inter- and motor-neurons that specify the ventral bias
of turns, and the \textsf{SMDV} are motor neurons that initiate
ventral turns.  We also see that states 6 (orange) and 5 (purple) often
produce a dip followed by a rise, respectively, of activation on cluster 8,
which together produce a dorsal turn.
%\todo{Need confirmation of these results.  Hard for me to tell difference between dorsal turn activation (orange to purple) and ventral turn (green to brown).}

@@

\paragraph{Input-driven transition probabilities}

\todo[inline]{Detail the model for how transition probabilities depend
  on oxygen and change in oxygen concentration.  Straightforward extension
  of the recurrent model above.}
